\vspace{-3em} \LARGE Hamiltonian Privilege

### 2023-02-24 Articles/2020-ClassicalMechanicsPhil/Submissions/ERKE/classical-phil-article-Erkenntnis.tex
Update classical-phil-article-Erkenntnis.tex
Edits from GC and JH in response to referee #2, comments 1 thru 4. Major changes in green font for ease of referee and editors. (Josh also changed a few words here and there for style).

@@

\title{\vspace{-3em} \LARGE Hamiltonian Privilege}			
%\author{\Large \textsuperscript{*\textdagger}}
\date{\vspace{-7.9ex} April 3rd, 2022}
\maketitle
%
\vspace{-8mm}

@@

Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding.\footnote{Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider a more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are \textit{not} categorically equivalent \parencites*[1182-83]{Barrett2}. Our construction of Hamiltonian mechanics in Section~\ref{derivation} matches this more general symplectic manifold formulation.} Whereas North's argument relies on a problematic criterion for comparing structure based on the size of symmetry groups \parencites[]{Swanson}, our argument avoids this pitfall. We show that a classical system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics, and this difference has interpretive consequences even within the hyper-regular domain. The construction at the heart of our argument provides an interpretation of classical mechanical systems in terms of their invariant phase space density. 
%We show that the invariant density fits more naturally within the Hamiltonian framework, providing another reason to privilege it over Lagrangian mechanics.
% Our formalism also provides a novel interpretation of the principle of stationary action within Hamiltonian mechanics.

Section~\ref{privileging} previews our argument for privileging Hamiltonian mechanics. We introduce three physical conditions that define classical systems while distinguishing Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require metaphysical commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. We then show that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. This completes our primary argument for privileging Hamiltonian mechanics. Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics.

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must provide a criterion for a framework being more fundamental than another. Of various notions of relative fundamentality, we focus on an entirely non-mysterious aspect. Given two formalisms that capture a class of systems, if one is logically more general, then it is more fundamental. This aspect of fundamentality is familiar from mathematics. A more fundamental framework encompasses a less fundamental one as a special case. It is in this sense that the generalized Stokes' theorem is more fundamental than well-known special cases, such as the fundamental theorem of calculus and Green's theorem. Similarly, an axiomatization which includes additional axioms is less fundamental than one which drops these axioms while maintaining the same expressive power. This aspect of fundamentality is distinct from aspects associated with differences in energy scale, such as how the Standard Model of particle physics is more fundamental than classical mechanics. We do not mean that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in this sense. As Gilton notes, ``many philosophers of science have moved away from thinking of fundamentality in science in terms of the very small toward thinking of it in terms of comprehensiveness, or breadth of applicability'' \parencites*[637]{Gilton2021}.
%we must provide a criterion for one framework being more or less fundamental than another
%\footnote{In Section~\ref{density}, we embrace a related non-metaphysical notion of fundamentality that pertains to physical entities rather than theoretical frameworks.}

To apply our criterion for relative fundamentality, we must show that Lagrangian mechanics is less general than Hamiltonian mechanics. We do this in Section~\ref{derivation} by recovering the mathematical formalisms of both Hamiltonian and Lagrangian mechanics from three physical conditions. In this section, we describe how our argument works in principle. Section~\ref{fundamentality} demonstrates some philosophical advantages of our criterion for relative fundamentality.\footnote{One might worry that our criterion is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where ``aboutness'' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality must address. It is not a special problem facing our criterion.} 
%We will show that the Lagrangian framework is less general because it requires an additional condition to hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.

%We will, however in Section 5 give a criterion for an entity to be physically fundamental
%These bolster our claim that this is an appropriate sense of fundamentality for philosophers to focus on.

@@

Shorn of many details, our argument goes as follows. First, we show that the formalism of Hamiltonian mechanics applies to systems that are both i) infinitesimally reducible and ii) follow a deterministic and reversible evolution. A system is \textit{infinitesimally reducible} provided that specifying the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Colloquially, we can say that the whole state equals the sum of its parts, where we take these infinitesimal parts to be classical particles. Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (i.e. we cannot specify the whole state by specifying the states of its subsystems).\footnote{The conditions of infinitesimal (ir)reducibility are similar to various notions of (non-)separability in the literature on physical holism \parencites[46, 124]{Healey}. We adopt our preferred terminology to avoid controversies surrounding how best to characterize separability.} We show in Section~\ref{infinitesimal} that the phase space of infinitesimally reducible systems must have at least the structure of a symplectic manifold. To recover Hamiltonian mechanics, it then suffices to identify when trajectories on phase space obey Hamiltonian evolution (i.e. evolve according to a symplectomorphism). Section~\ref{deterministic} shows that this requires the system to follow a \textit{deterministic and reversible evolution}, i.e. the present state of the system uniquely determines both its future and past states. We thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution.  
%Since we won't have to settle controversies surrounding these terms, we adopt our preferred terminology.

Next, we note that some classical systems satisfy a third, further condition, which we call \textit{motion equivalence}. Motion equivalence holds when the system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics (Section~\ref{motion}). As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics.

@@

% % optional: address refer 1 concern about the clarity of this argument/example: ``what if momentum is defined in terms of velocity? Or is the idea to know that we also need to know the mass?"

Of course, different definitions of ``Hamiltonian'' and ``Lagrangian'' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems also satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they lie in the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems. For details of a Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the details of our argument in Section~\ref{derivation}, we describe how we avoid central problems facing North's privileging of Hamiltonian mechanics. 
%, although Curiel seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.

@@

Our derivation has three main steps. Section~\ref{infinitesimal} shows that the phase space of infinitesimally reducible systems must have at least the structure of a symplectic manifold. For systems with deterministic and reversible evolution, Section~\ref{deterministic} shows they must evolve according to a symplectomorphism. Hence, we derive the formalism of Hamiltonian mechanics from the conditions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics.

@@

\noindent
This assumption characterizes classical systems as those whose infinitesimal parts completely specify the state of the system. We define a \textit{classical particle} as one of these infinitesimal parts, namely the limit of recursive subdivision.\footnote{This definition of a classical particle differs from viewing point-particles as extensionless \parencites[]{Butterfieldpoints} or zero-dimensional \parencites[]{Wilson}.}
% % % For further discussion, see \textcites{shpmp}. [omit this for review].

@@

When the system is characterized by continuous quantities, $\mathcal{S}$ is a manifold. For $U \subseteq \mathcal{S}$, we define $\mu(U) := \int_U \rho_c(s) d\mathcal{S}$ to measure the number of particles in region $U$. \textit{State variables} $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called ``coordinates'' and $\xi$ is a coordinate map. To avoid confusion, we will call the coordinates of the state space $\mathcal{S}$ ``state variables.'' As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument. It allows us to compare and track the amount of matter in different possible states.

@@

\noindent In our case, this applies to both the state of the whole system and the state of its parts.

@@

Intuitively, our argument is the inverse of Liouville's theorem: instead of positing Hamiltonian evolution and finding conservation of areas and volumes, deterministic and reversible evolution imposes the conservation of areas and volumes (i.e. sets of states are mapped to sets of states of equal size). This leads to Hamiltonian evolution. In summary, we have shown that any system that satisfies the conditions of both infinitesimal reducibility and deterministic/reversible evolution is a classical Hamiltonian system.
% [(ref 1): expand on how our argument relates to standard proofs of Liouville's theorem. Is it fair to say that we are privileging Hamiltonian mechanics because it forms the basis for a proof of Liouville's theorem?]

@@

\noindent If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space.

@@

In this case---which we call \textit{weak equivalence}---$v^i(q^j, p_k)$ is invertible at every $q^j$. Therefore, for fixed $q$, the Jacobian matrix  $\frac{\partial v^i}{\partial p_k}$ must be invertible. Let $H$ denote the Hamiltonian. By Hamilton's equations, $v^i = d_t q^i = \frac{\partial H}{\partial p_i}$, so the Hessian $\frac{\partial^2 H}{\partial p_k \partial p_i} = \frac{\partial}{\partial p_k} (\frac{\partial H}{\partial p_i})$ equals the Jacobian $\frac{\partial v^i}{\partial p_k}$. Since the Jacobian is invertible, the Hessian must be non-zero everywhere, and therefore it must have the same sign everywhere---which we take to be positive. Weak equivalence allows us to express particle states in terms of either position/momentum or position/velocity variables. These are exactly the cases where a Lagrangian can be constructed, and where the Lagrangian leads to a unique solution (via the stationary action principle). Hence, weak equivalence suffices for the existence of a hyper-regular Lagrangian. As discussed in Section~\ref{Curiel}, Lagrangian mechanics only yields a unique evolution in the hyper-regular domain.\footnote{Weak equivalence corresponds to a general treatment of Lagrangian mechanics that does not presuppose Riemannian geometry. As we discuss in Section~\ref{Curiel}, Equation~\ref{weak_equivalence} corresponds to the kinematical constraints central to Curiel's \parencites*[]{Curiel} argument for privileging Lagrangian mechanics.}
%; otherwise, the positions and velocities do not determine how the system evolves

>>>>

\usepackage[dvipsnames]{xcolor}%lets us define a new color 
\definecolor{greeo}{RGB}{91,173,69}
\colorlet{green}{greeo} %redefines color `green' to color `greeo' (or any other color)

@@

\title{\vspace{-3em} \LARGE Hamiltonian Privilege}			
%\author{\Large \textsuperscript{*\textdagger}}
\date{\vspace{-7.9ex} \today}
\maketitle
%
\vspace{-8mm}

@@

Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding.\footnote{Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider a more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are \textit{not} categorically equivalent \parencites*[1182-83]{Barrett2}. Our construction of Hamiltonian mechanics in Section~\ref{derivation} matches this more general symplectic manifold formulation{\color{green} and illuminates why it is more general.}} Whereas North's argument relies on a problematic criterion for comparing structure based on the size of symmetry groups \parencites[]{Swanson}, our argument avoids this pitfall. We show that a classical system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics. {\color{green} Unlike Lagrangian mechanics, Hamiltonian mechanics provides a unique solution to the equations of motion even outside the hyper-regular domain. Hence, Lagrangian mechanics is restricted to the hyper-regular regime.} The construction at the heart of our argument provides a novel interpretation of classical mechanical systems in terms of their invariant phase space density.

Section~\ref{privileging} previews our argument for privileging Hamiltonian mechanics. We introduce three physical conditions that define classical systems while distinguishing Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require metaphysical commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. {\color{green} Our derivation shows that even non-hyper-regular Hamiltonian systems are physically interpretable.} We then show that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. This completes our primary argument for privileging Hamiltonian mechanics. Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics.

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must provide a criterion for assessing whether one framework is more fundamental than another. Of various notions of relative fundamentality, we focus on an entirely non-mysterious one. Given two formalisms that capture a class of systems, if one is logically more general, then it is more fundamental. This aspect of fundamentality is familiar from mathematics. A more fundamental framework encompasses a less fundamental one as a special case. It is in this sense that the generalized Stokes' theorem is more fundamental than special cases such as the fundamental theorem of calculus and Green's theorem. Similarly, an axiomatization that includes more axioms is less fundamental than one which drops these axioms while maintaining the same expressive power.\footnote{One might worry that our criterion is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where ``aboutness'' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality must address. It is not a special problem facing our criterion.}  This aspect of fundamentality is distinct from aspects associated with differences in energy scale, such as how the Standard Model of particle physics is more fundamental than classical mechanics. We do not mean that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in this sense. As Gilton notes, ``many philosophers of science have moved away from thinking of fundamentality in science in terms of the very small toward thinking of it in terms of comprehensiveness, or breadth of applicability'' \parencites*[637]{Gilton2021}.
%we must provide a criterion for one framework being more or less fundamental than another
%\footnote{In Section~\ref{density}, we embrace a related non-metaphysical notion of fundamentality that pertains to physical entities rather than theoretical frameworks.}

To apply our criterion for relative fundamentality, we must show that Lagrangian mechanics is less general than Hamiltonian mechanics. We do this in Section~\ref{derivation} by recovering the mathematical formalisms of both Hamiltonian and Lagrangian mechanics from three physical conditions. {\color{green}Our derivation illuminates the relationship between Hamiltonian and Lagrangian mechanics. As North \parencites*[]{North} argues, there is a sense in which Hamiltonian mechanics is more general than Lagrangian mechanics. Barrett \parencites*[]{Barrett2} makes this sense more precise by showing that the symplectic manifold formulation of Hamiltonian mechanics is categorically inequivalent to Lagrangian mechanics formulated on a tangent bundle. Still, one might wonder why it is that Hamiltonian mechanics can be formulated in a more general framework than Lagrangian mechanics. Additionally, one might wonder whether Lagrangian mechanics could be meaningfully formulated in an analogously general manner. Our derivation helps answer these questions: the physical principles that undergird the symplectic manifold formulation of Hamiltonian mechanics are simply more general than the principles required for Lagrangian mechanics.}\footnote{{\color{green}We thank a referee for this journal for this helpful way of framing the significance of our derivation.}} In this section, we describe how our argument works in principle. Section~\ref{fundamentality} demonstrates some philosophical advantages of our criterion for relative fundamentality.
%We will, however in Section 5 give a criterion for an entity to be physically fundamental
%These bolster our claim that this is an appropriate sense of fundamentality for philosophers to focus on.

@@

Shorn of many details, our argument goes as follows. First, we show that the formalism of Hamiltonian mechanics applies to systems that are both i) infinitesimally reducible and ii) follow a deterministic and reversible evolution. A system is \textit{infinitesimally reducible} provided that specifying the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Colloquially, we can say that the whole state equals the sum of its parts, where we take these infinitesimal parts to be classical particles. Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (i.e. we cannot specify the whole state by specifying the states of its subsystems).\footnote{The conditions of infinitesimal (ir)reducibility are similar to various notions of (non-)separability in the literature on physical holism \parencites[46, 124]{Healey}. We adopt our preferred terminology to avoid controversies surrounding how best to characterize separability.} We show in Section~\ref{infinitesimal} that {\color{green} if one chooses to model continuous quantities using a manifold, then this manifold must be symplectic.} To recover Hamiltonian mechanics, it then suffices to identify when trajectories on phase space obey Hamiltonian evolution (i.e. evolve according to a symplectomorphism). Section~\ref{deterministic} shows that this requires the system to follow a \textit{deterministic and reversible evolution}, i.e. the present state of the system uniquely determines both its future and past states. We thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution. {\color{green} Importantly, our derivation encompasses Hamiltonian mechanics in the non-hyper-regular regime as well.}

Next, we note that some classical systems satisfy a third, further condition, which we call \textit{motion equivalence}. Motion equivalence holds when a system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics (Section~\ref{motion}). {\color{green} Since a system satisfies motion equivalence if and only if it is hyper-regular, this condition illuminates the physical significance of hyper-regularity. Moreover, since Lagrangians lack unique solutions to the equations of motion outside the hyper-regular regime, we take ourselves to have recovered the physically significant Lagrangian systems.} As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics.

@@

Of course, different definitions of ``Hamiltonian'' and ``Lagrangian'' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems also satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they lie in the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems. For details of a Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the details of our argument in Section~\ref{derivation}, we describe how we avoid central problems that arise in North's privileging of Hamiltonian mechanics. 
%, although Curiel seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.

@@

Our derivation has three main steps. Section~\ref{infinitesimal} shows that {\color{green} if one represents continuous quantities using a manifold, then this manifold must be symplectic.} For systems with deterministic and reversible evolution, Section~\ref{deterministic} shows they must evolve according to a symplectomorphism. Hence, we derive the formalism of Hamiltonian mechanics from the conditions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics. {\color{green}In fact, when motion equivalence fails, Lagrangian mechanics does not yield unique solutions to the equations of motion.}

@@

\noindent
This assumption characterizes classical systems as those whose infinitesimal parts completely specify the state of the system. We define a \textit{classical particle} as one of these infinitesimal parts, namely the limit of recursive subdivision.\footnote{This definition of a classical particle differs from viewing point-particles as extensionless \parencites[]{Butterfieldpoints} or zero-dimensional \parencites[]{Wilson}.} {\color{green} Ultimately, this condition entails that we can describe the state of the whole system as a statistical distribution over the state space of its parts. If we model the state space of the parts using a manifold, this manifold must be differentiable so that it can support a density at each point. Moreover, given that this density must be a function of the state, and not of the coordinates, we show that the manifold must be symplectic.}
% % Could add: readers who are satisfied with this overview can proceed to the next subsection
% % % For further discussion, see \textcites{shpmp}. [omit this for review].

@@

When the system is characterized by continuous quantities, {\color{green}$\mathcal{S}$ can be taken to be a manifold}.\footnote{{\color{green}The use of manifolds is, technically, a choice of how the physics is represented. One could make another choice, for example, by representing the state variables not as local coordinates but as an algebra of operators that act on the state. In this case, we conjecture that a similar argument would demonstrate an analogue of a symplectic structure for the different mathematical representation.}} For $U \subseteq \mathcal{S}$, we define $\mu(U) := \int_U \rho_c(s) d\mathcal{S}$ to measure the number of particles in region $U$. \textit{State variables} $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called ``coordinates'' and $\xi$ is a coordinate map. To avoid confusion, we will call the coordinates of the state space $\mathcal{S}$ ``state variables.'' As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument. It allows us to compare and track the amount of matter in different possible states.

@@

\noindent In our case, this applies to both the state of the whole system and the state of its parts. {\color{green} Under this assumption, we show that the symplectic structure must be preserved over time, which gives us Hamiltonian evolution.}

@@

Intuitively, our argument is the inverse of Liouville's theorem: instead of positing Hamiltonian evolution and finding conservation of areas and volumes, deterministic and reversible evolution imposes the conservation of areas and volumes (i.e. sets of states are mapped to sets of states of equal size). This leads to Hamiltonian evolution. In summary, we have shown that any system that satisfies the conditions of both infinitesimal reducibility and deterministic/reversible evolution is a classical Hamiltonian system. {\color{green} Note in particular that our derivation is not restricted to only hyper-regular Hamiltonians. As we will see, this is a key difference between Hamiltonian and Lagrangian mechanics.}

%Since, the only requirement that we have imposed on the Hamiltonian is that the equations yield a unique solution, our derivation applies to Hamiltonians that are not hyper-regular. }

% since we have not imposed hyper-regularity,   }
% [(ref 1): expand on how our argument relates to standard proofs of Liouville's theorem. Is it fair to say that we are privileging Hamiltonian mechanics because it forms the basis for a proof of Liouville's theorem?]

@@

\noindent If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space. {\color{green}This assumption means that momentum can be expressed as a function of position and velocity, which mathematically corresponds to the case in which the Hamiltonian is hyper-regular. In this regime we find Lagrangian mechanics. Moreover, to be able to express the symplectic form, and therefore the density, over position and velocity, we find that momentum must be a linear function in velocity, which constrains the Hamiltonian to be the one of a massive particle under scalar and vector potential forces.}

@@

In this case---which we call \textit{weak equivalence}---$v^i(q^j, p_k)$ is invertible at every $q^j$. Therefore, for fixed $q$, the Jacobian matrix  $\frac{\partial v^i}{\partial p_k}$ must be invertible. Let $H$ denote the Hamiltonian. By Hamilton's equations, $v^i = d_t q^i = \frac{\partial H}{\partial p_i}$, so the Hessian $\frac{\partial^2 H}{\partial p_k \partial p_i} = \frac{\partial}{\partial p_k} (\frac{\partial H}{\partial p_i})$ equals the Jacobian $\frac{\partial v^i}{\partial p_k}$. Since the Jacobian is invertible, the Hessian must be non-zero everywhere, and therefore it must have the same sign everywhere---which we take to be positive. {\color{green} This, by definition, means that the Hamiltonian is hyper-regular if and only if weak equivalence is satisfied.} Weak equivalence allows us to express particle states in terms of either position/momentum or position/velocity variables. These are exactly the cases where a Lagrangian can be constructed, and where the Lagrangian leads to a unique solution (via the stationary action principle). Hence, weak equivalence suffices for the existence of a hyper-regular Lagrangian. As discussed in Section~\ref{Curiel}, Lagrangian mechanics only yields a unique evolution in the hyper-regular domain.\footnote{Weak equivalence corresponds to a general treatment of Lagrangian mechanics that does not presuppose Riemannian geometry. As we discuss in Section~\ref{Curiel}, Equation~\ref{weak_equivalence} corresponds to the kinematical constraints central to Curiel's \parencites*[]{Curiel} argument for privileging Lagrangian mechanics.}
%; otherwise, the positions and velocities do not determine how the system evolves
